Spring haul-out behavior of seals in the Bering and Chukchi seas

### 2023-09-15 analysis/paper/Letter_Response_to_Reviewers.qmd

---
title: " "
subtitle: Maunscript 2022:04:72978
format: 
  pdf:
    include-in-header:
      text: |
        \usepackage[dvipsnames]{xcolor}
filters:
   - latex-environment
environments: [author-response]
editor: visual
---

```{=tex}
\newenvironment{author-response}{\color{Bittersweet}}

```
Josh M. London, PhD\newline
7600 Sand Point Way NE\newline Seattle, Washington, 98115\newline josh.london\@noaa.gov\newline

**Manuscript 2022:04:72978**

\hfill \break

Dr. Prange,

First, please accept my apologies for the significant delay in returning revisions on this manuscript. A number of life and work priorities interfered with a more timely response schedule and I sincerely appreciate the flexibility *PeerJ* extends to authors. We are commited to a quick turnaround on any future revisions or tasks associated with this manuscript and we look forward to a final publication in *PeerJ*.

We feel, while many of the comments and suggestions from reviewers were minor, they deserved appropriate time, consideration, and exploration. We hope our responses here reflect that consideration and we are grateful to the reviewers for their thoughtful comments. Lastly, we also used this opportunity to improve the quality of the underlying telemetry data set and expand the reproducibility of the analysis to meet or exceed expectations for open science, data availability, and code availability.

Below, please find detailed responses to each of the reviewer comments. In addition to this letter, several of the suggestions are documented as specific issues within the GitHub repository and the full suite of updates are all contained and documented within a single pull request. We have also provided a pdf document that indicates all changes between the revised manuscript and the submitted manuscript.

Thank you, again, for your time and consideration. And, additionally, thank you to the reviewers for their time. It is much appreciated and we feel the manuscript is a better product as a result.

Regards

Josh M London, PhD

\pagebreak

# Response to Reviewers

Here, we respond directly to each of the suggested edits and comments. These responses are organized by reviewer. All responses are prefaced with **\[author response\]** and line numbers in the original submission are used as reference. \hfill \break

\hrule

## **Reviewer 1 (A. Fishbach)**

> *The authors have put substantial efforts into setting the context for their analysis, citing relevant studies. I would like to see a more direct approach in the framing of their problem: the improvements of abundance estimates required by the Marine Mammal Protection Act and Endangered Species Act. I would like to see a specific example of how the modeled availability would have affect population abundance estimates by apply the availability model to aerial survey data and providing an adjusted abundance estimate.*

::: author-response
**\[author response\]** We address this in the manuscript by introducing the importance of the statutory requirements of the MMPA and ESA early in the second paragraph of the introduction. We hope this frames the important link in the mind of the reader between the work described here and the need for improved abundance estimates.

Regarding the desire for specific examples of how the modeled availability would have impacted population abundance estimates from aerial survey data, we don't have the ability to provide such an example because the aerial survey data are not yet published and available. Such an example and discussion might be better suited for subsequent papers that focus on presentation of the aerial survey data and the abundance estimation process.
:::

> *The findings are appear robust and intuitive. The authors need to improve the presentation of the findings to improve clarity and help the reader understand how they addressed the knowledge gap.*

::: author-response
**\[author response\]** The reviewer provided numerous comments and suggestions annotated throughout the PDF that address the concern of presentation and clarity. In the next section we provide responses to those annotations as best as possible
:::

### **Reviewer 1 (A. Fishbach) - PDF Annotated Comments & Suggested Edits**

> ***Title:** The reviewer suggests removing and aerial survey detectability from the title of the manuscript and suggests adding the seasonal time period our study focused on.*

::: author-response
**\[author response\]** We have changed the title to: Spring haul-out behavior of seals in the Bering and Chukchi seas
:::

> ***Abstract:** Can you make a statement in your abstract that use of modeled haulout behavior using covariates available from regional reananalysis data improved the confidence intervals on availability estimates and that this would improve aerial survey abundance estimation efforts?*

::: author-response
**\[author response\]** We don't feel comfortable making a precision comparison, simply because there are too many differences between the analyses, including (1) the data sets are different (the present study has substantially more records), and (2) the model structures are different (the present study has more parameters). As such, it will be difficult to ascribe a specific reason to precision increases or decreases.
:::

> *Please use direct language. You are not making findings on how seals rely on sea ice, rather you are improving methods to estimate their availability for improved aerial survey population estimation efforts. Here you are focused on modeling ice sea availability to aerial survey so that aerial survey counts may be appropriately interpreted. Please be clear that you are considering ice seal behaviors that may be telemtered as "dry out of water behaviors" that may be interpreted as behaviors that place the seal on top of the ice and available for observation by an aerial survey.*

::: author-response
**\[author response\]** The opening sentence of the abstract is a statement regarding the reliance of ice-associated seals on sea-ice for a range of important life history behaviors. We think this statement is important introduction for readers who are not familiar with ice-associated seals. We do agree with the reviewer's suggestion for more direct language and have made edits to the beginning of the abstract with this in mind.
:::

> ***Abstract:** Did you use previously published data, or did you collect it?  Be specific using an active voice so that readers may understand your effort.*
>
> *For better flow, please indicate the behavioral data logging time scale here when you describe the data collection, not later when you speak to the analytical construct.*

::: author-response
**\[author response\]** We have edited this to simply state *data from satellite-linked bio-loggers* ... we have also altered this sentence to specifically mention the hourly timescale for the bio-logger data
:::

> *The term "basking" implies a subset of activities that ice seals may engage in when hauled out on sea ice and not an activity that can be directly telemetered using your data logger.  Yet, what you are describing is "haul-out behavior".  Please use this term here.*

::: author-response
**\[author response\]** edited to remove use of 'basking'
:::

> *Can you help your naive reader understand why you consider these covariates to be of interest and why others were not considered.*

::: author-response
**\[author response\]** The main body of the paper includes this discussion. Given the limited space and purpose of the abstract we feel it best to simply list the covariates of interest
:::

> ***Line 41:** Please use specific terms and not metaphors so that second-language English readers may better understand your writing.*

::: author-response
**\[author response\]** We have edited 'ripple effects' to 'tangible effects'
:::

> ***Line 41:** What about human communities that don't live in the region, but that send large LNG tankers and other vessel through the region?  The loss of sea ice has effects on those communities and activities too.*

::: author-response
**\[author response\]** There is a seemingly endless list of impacts of loss of sea ice to human communities (and ecosystems) beyond the Arctic region to the rest of the world. However, for this sentence, we wanted to keep the reader focused on the Arctic region.
:::

> ***Line 44:** Please use direct language you mean biologically important functions?*

::: author-response
**\[author response\]** The term 'life history function' is a commonly used, direct term.

Here is a definition from Oxford:

"The life history of an organism is its pattern of survival and reproduction, along with the traits that directly affect survival and the timing or amount of reproduction. Rates of survival and reproduction can be estimated across age classes, or across different stages in organisms with complex life cycles. Life history traits include growth rate; age and size at sexual maturity; the temporal pattern or schedule of reproduction; the number, size, and sex ratio of offspring; the distribution of intrinsic or extrinsic mortality rates (e.g., patterns of senescence); and patterns of dormancy and dispersal."

To ensure the reader is clearly aware of which life-history behaviors/functions we are referring to we provide specific examples within the sentence.
:::

> ***Line 50:** What does "readiness" mean in the context of a marine mammal?*

::: author-response
**\[author response\]** This is a good catch and we have edited the sentence to, now, read:

"Knowledge about evolutionary constraints on the timing of reproductive and molting behavior is generally lacking, so it is difficult to predict how or if ice-associated seal species might adapt to future changes (e.g., by adjusting pupping or molting schedules to earlier dates or different locales)"
:::

> ***Line 56:** Perhaps you mean to offer a topic sentence, stating "How ice-assoicated seals use sea ice as a haulout platform varies betweens species".*

::: author-response
**\[author response\]** We have adopted this suggestion
:::

> ***Line 66-68:** There is so much to state about ice-associated seals in the Pacific Arctic that this reference to Antarctic work seems out of place here.*

::: author-response
**\[author response\]** We have moved this reference to later in the paragraph so the first references are about Arctic work. While the example is from the Antarctic, it is an important, early example of analysis using bio-logger data to estimate haul-out behavior in seals.
:::

> ***Line 77:** suggested edit*

::: author-response
**\[author response\]** We have adopted the suggested edit
:::

> ***Line 84:** Do you mean age-class? The term life-history implies knowledge of previous behavioral efforts or occupancy of regions of interest.   Here you are specifically interested in age-class.*

::: author-response
**\[author response\]** We have edited to use the term 'demographic'
:::

> ***Line 92:** Please help your reader understand why are you doing your analysis and not offer a wide and jargon rich inventory of concepts that may be considered.*

> ***Line 93-94:** Please keep your introduction focused on establishing the knowledge gap that you are seeking to fill. Please do not lead your reader to the numerous other knowledge gaps that are not being addressed by your current study.*

::: author-response
**\[author response\]** We have removed the sentence in question
:::

> ***Line 99:** Please use the term "hauled-out". Neither the biologger nor the aerial survey is able to distinguish among the various seal behaviors that result in them being hauled out and available for view by an aerial survey.*

::: author-response
**\[author response\]** noted, and the terms "hauled out", "to haul out", etc is used consistently throughout the manuscript
:::

> ***Line 114:** You have indicate that you would consider sea ice extent. Here you introduce the concept of icescape. Please introduce this concept earlier and indicate how you will incorporate this into your analysis.*

::: author-response
**\[author response\]** changed "icescape" to "sea-ice extent"
:::

> ***Line 117:** Please use an active voice for clarity.*

::: author-response
**\[author response\]** This sentence has been changed to:

"For this study we used haul-out behavior data and location estimates from bio-loggers deployed on bearded, ribbon, and spotted seals in the Bering, Chukchi, and western Beaufort seas by multiple organizations as part of collaborative investigations from 2005 through 2020."
:::

> ***Line 120:** You summarize the years during which land-based studies applied bio-loggers.\
> Please do the same for the offshore deployments.*

::: author-response
**\[author response\]** This sentence now reads:

"Ship-based capture events occurred during spring near the southern ice edge in the Bering Sea between 2005 and 2018."
:::

> ***Line 124:** Please provide the geographic context of your tag deployments and data collection using a plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 2). Please consider using a tabulation that enables readers to understand the geographic context of the deployments similar to that used by Jay et al. 2012, table 1.*

::: author-response
**\[author response\]** We feel the spatial distribution of the actual transmitted data -- compared to the deployment location -- is most informative and relevant for the audience's understanding of how our data are distributed spatially and how well data represents the range of each species. We provide this visually in Figure 3 and adding an additional layer to that figure to show deployments would be an additional complication. To ensure the audience has a clear understanding of our study area, we have added an additional figure with a map of the Bering and Chukchi sea region that indicates generally where deployments were initiated.
:::

> ***Line 126:** Please use active voice for all actions that you took in your data collection, preparation and analysis.*

::: author-response
**\[author response\]** Noted and addressed
:::

> ***Line 129-135:** Please move this to the discussion. This is a summary of your data findings, not a description of your methods.*

::: author-response
**\[author response\]** We can see the reviewer's point. However, we do not feel this topic is a summary of our data findings in this paper. Instead, our view is that this information provides the context and rationale for our methodological choice regarding bio-logger type and attachment methods. It is also important regarding the analytic methods -- preference for flipper tag haul-out data over those from a tag adhered to the dorsal/head. If we were to move these important points to the discussion, we worry those points would be left disconnected.
:::

> ***Line 136:** This is a discussion of your methods. This is not your methods. Please state how the data was collected in your Methods section and discuss how data collection have been uncertain or biased in your discussion section.*

> ***Line 141-143:** Please move to your discussion.*

::: author-response
**\[author response\]** We appreciate this suggestion from the reviewer and we have moved large portions of this paragraph to the discussion. However, some information regarding age/sex class determination is left here so the reader has context for Table 1.
:::

> ***Line 144:** Please indicate how this was determined during data collection.*

::: author-response
**\[author response\]** We have edited this sentence some to clearly communicate that field determination of reproductive status (and, thus ideal delineation of sub-adult and adult age class) was not possible.
:::

> ***Line 147-148:** Please allow your readers the opportunity to understand the geographic context of these deployments through use of symbols that maybe used to cross-reference the deployment locations in a map that indicates deployment locations along with filtered movement tracks.*

::: author-response
**\[author response\]** As previously addressed, there are now two figures available that provide geographic context of the deployments. Figure 3 shows the geographic distribution of the data from the deployments. The paper is not presenting or discussing movement ecology and, therefore, we do not present filtered movement tracks. We keep the focus on haul-out behavior data.
:::

> ***Figure 1:** Please consider bar plot similar to that used by Jay et al. (2012: 10.3354/meps10057 figure 3), replacing the region with the seal species.*

::: author-response
**\[author response\]** This figure has been updated with this suggestion in mind
:::

> ***Line 151:** Please state clearly how you determined the end of a deployment. Here you indicate various elements of the haulout chronology, movement and dive chronology that may have been used, however not reader could replicate your methods base don this text.*

::: author-response
**\[author response\]** We've added the following text to the paragraph:

"For example, a data record that ends with several consecutive days (\~10+ days) of 100% dry observations and with locations indicating the tag was on land would be truncated to not include the final stretch of 100% dry observations. The vast majority of deployments end with the device detaching in the water and the deployment end date is obvious. There is no perfect algorithm for identifying deployment end dates and each deployment in question must be considered separately. While not perfect, we are confident our reliance on expert opinion and examination of multiple data streams provides the best option."
:::

> ***Line 152:** Deployment activity is known to disrupt normal pinniped behavior. Did you censure some period of data collection during and after the deployment to ensure that that disruptions from the deployment activity did not bias your analysis?*

::: author-response
**\[author response\]** For these phocid species, we have no empirical data (or even anecdotal observational data) on how captures may or may not disrupt 'normal' behavior. This is likely to be influenced by the capture method, how long the animal was restrained, whether the animal was with a dependent pup, whether the animal was molting, environmental variables such as wind and temperature, etc etc. Given all these unknowns, we have not applied a censure to the beginning of our deployment data. We can't 100% ensure this choice doesn't result in some level of bias. However, given the large sample size and duration of deployments we expect any bias imposed is negligible.
:::

> ***Line 154:** This is a result. Please indicate the method that you used to distinguish "plausible" from "implausible" telemetered seal behavior. In your Results section, you may indicate how method was able to identify occasions of implausible seal behavior and any patterns that arose relative to tag model or attachment.*

::: author-response
**\[author response\]** This sentence has been moved to the results section. Additionally, we've provided a brief explanation of implausible along with a reference for more details. For these devices, the percent-dry timeline reported 100% dry for several weeks followed by an additional period of time that was more typical of expected bearded seal behavior. During the extended dry time the locations indicated movement of the tag that would not be simply drifting ice. We also observed increasing transmission rates. The working hypothesis (i.e. unconfirmed guess) is that the wet/dry sensor surface became obscured by some corrosive element such that when in the water conductivity was prevented and, thus, recorded as dry.
:::

> ***Line 156-164:** Please move to the Results section. Please move the discussion of how molt timing affects termination of tag deployments to the discussion section.*

::: author-response
**\[author response\]** This figure has been moved to the results section along with additional descriptive text. We have left the discussion regarding molt timing and termination of the data record here to better communicate the point to the reader and prevent an isolated mention in the discussion that has no anchor.
:::

> ***Line 169-170:** Thus, the a successful transmission of data was required at no greater than 48 h intervals.*

::: author-response
**\[author response\]** We're not sure how to interpret this comment. The main point of the sentence is to explain the rationale/benefits of transmitting the limited selection of percent dry values. We have expanded the text to provide more details regarding the caching of messages and data transmission.
:::

> ***Line 174:** Do you mean "atmospheric" magnetic interference.\
> Presumably "extra-terrestrial atmospheric" events (such as events on Venus or Mars) played no role in the transmission of the data signal to the Argos receivers in orbit around Earth.*

::: author-response
**\[author response\]** We were using a less strict definition of extra-terrestrial to refer to interference that wasn't Earth/surface-based. We've edited to 'atmospheric interference'
:::

> ***Line 174-175:** This note belongs in the Discussion, not the Results.*

::: author-response
**\[author response\]** We feel this note is important to acknowledge in the methods because it speaks to the rationale for our methodological choices (bio-logger type, programming choices, analytical choices).
:::

> ***Figure 3:** Do you mean that each cell "encompasses" and area of 50 km\^2?*

::: author-response
**\[author response\]** We have edited to the reviewer's suggested word 'encompasses'
:::

> ***Line 197:** Please use the active voice.*

::: author-response
**\[author response\]** Thank you for catching this. We have edited this sentence to active voice
:::

> ***Line 198-199:** For most tags, your daily locations are from the Argos system, not from your data loggers. However, you included some FastLoc tag deployments. Please re-write for clarity of what these location are.*

::: author-response
**\[author response\]** The previous paragraph describes how we determined daily locations. We see the reviewers point re: the data not technicaly coming from our data loggers. We have edited the sentence for additional clarity.
:::

> ***Line 199:** What is a "key" weather variable? Please write for clarity.*

::: author-response
**\[author response\]** We've edited this sentence and removed the word 'key'
:::

> ***Line 203:** This is a procedural detail. Other readers may replicate your method without needing to do this. Please omit this detail.*

> ***Line 207-208:** Readers can replicate your methods without the details on converting the NARR wind vectors into wind speed.*

::: author-response
**\[author response\]** We're not sure why the reviewer would recommend omitting a 'procedural detail' as part of our methods. We feel it is important to clearly communicate our procedure for processing the large NARR data set and that we used the native grid size of 32km. Regarding the calculation of wind vectors, we include this detail because many readers, like us initially, may not be well versed in how wind vectors components are converted to wind speed.
:::

> ***Line 210:** Please use an active voice: "For all seal species we considered the following variables when modeling the hourly haulout behavior..."*

::: author-response
**\[author response\]** We have edited this sentence with the reviewer's suggestion
:::

> ***Line 224:** Using the term "major focus" implies that there may be numerous minor foci. Can you not state that this is the goal of the study? Ancillary natural history insights can be discussed incidental to this, but are not driving the research. The statutory requirements of the MMPA and ESA are driving your study. These Federal laws require robust methods to monitor population status. Your study seeks to build this method. Discussion of natural history belong in your Discussion section and not in the framing of your study. Please help the reader focus on why the Federal government is doing this study.*

### 2023-09-15 analysis/paper/London_HauloutBeahvior.Rmd
new figure/updates for deployment locs
addresses #50

theme_set(theme_minimal(base_family = 'Source Sans Pro'))
theme_replace(plot.title.position = "plot")
```

@@

(ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020. Colored regions shown are minimum concave polygons buffered by 60 km for enhanced visibility. See Supplemental Material S1 for specific details. Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis. See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study.

```{r deployplot, include=TRUE, fig.asp = 0.45, fig.cap = "(ref:deployMapCaption)"}
targets::tar_load(deploy_details, store = here::here("_targets"))

@@

@@

p <- ggplot() +
  layer_spatial(deploy_sf, aes(fill=species),
                linewidth=0) +
  annotation_spatial(world,
                     linewidth=0,
                     fill = "grey80") +
  facet_grid(cols = vars(species), labeller = species_labels,
             switch="x") +
  scale_fill_manual(values = met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
                    aesthetics = "fill", 
                    name = "species") + 
  ggtitle("Areas of initial deployment location",
          subtitle = 
                 paste0("Bio-loggers were initially deployed on ",
                "<span style = 'color:",species_colors['Bearded seal'],";'>**bearded**, </span> ", 
                "<span style = 'color:",species_colors['Ribbon seal'],";'>**ribbon**, </span> ",  
                "and ", 
                "<span style = 'color:",species_colors['Spotted seal'],";'>**spotted**</span> seals ", 
                "across multiple regions within the Bering, Chukchi, and western Beaufort seas.")) +
  theme(legend.position = "none",
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_markdown(face="bold"),
        plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))

>>>>

theme_set(theme_minimal(base_family = 'Source Sans Pro',
                        base_size = 9))
theme_replace(plot.title.position = "plot")
```

@@

(ref:deployMapCaption) Initial deployment areas for each seal species (bearded, ribbon, and spotted seals) in the study between 2005 and 2020. Regions shown for each species are minimum concave polygons buffered by 60 km for enhanced visibility. The larger region (shaded yellow) indicates the spatial extent of aerial survey effort to date in US waters of the Bering and Chukchi seas. Deployments were initiated across a range of months but only data from 1 March to 15 July were included in the analysis. See Figure \@ref(fig:dataMap) for the spatial distribution of observed data used in the study and Supplemental Materials 1 for additional deployment details.

```{r deployplot, include=TRUE, fig.asp = 0.45, fig.cap = "(ref:deployMapCaption)"}
targets::tar_load(deploy_details, store = here::here("_targets"))
targets::tar_load(survey_area, store = here::here("_targets"))

@@

survey_area <- survey_area %>% 
  sf::st_transform(st_crs(deploy_sf))

@@

label_coords_bering <- st_sfc(st_point(c(178.5,57)), crs = 4326) %>%
  st_transform(st_crs(deploy_sf))
label_coords_chukchi <- st_sfc(st_point(c(-176,73)), crs = 4326) %>%
  st_transform(st_crs(deploy_sf))
label_coords_beaufort <- st_sfc(st_point(c(-146.5,72.25)), crs = 4326) %>%
  st_transform(st_crs(deploy_sf))

p <- ggplot() +
  layer_spatial(survey_area, color = "#ddc000", 
                fill = "#ddc000", 
                alpha = 0.15,
                linewidth = 0.5) +
  layer_spatial(deploy_sf, aes(fill=species),
                linewidth=0) +
  annotation_spatial(world,
                     linewidth=0,
                     fill = "grey80") +
  geom_sf_text(data = label_coords_bering, aes(label = "Bering sea"), size = 1.75, color = "grey40") +
  geom_sf_text(data = label_coords_chukchi, aes(label = "Chuckchi \nsea"), size = 1.75, color = "grey40") +
  geom_sf_text(data = label_coords_beaufort, aes(label = "Beaufort \nsea"), size = 1.75, color = "grey40") +
  facet_grid(cols = vars(species), labeller = species_labels,
             switch="x") +
  scale_fill_manual(values = met.brewer("Isfahan2",n=3,direction=-1,override.order = TRUE),
                    aesthetics = "fill", 
                    name = "species") + 
  ggtitle("Initial bio-logger deployment locations by species",
          subtitle = 
                 paste0("Bio-loggers were initially deployed on ",
                "<span style = 'color:",species_colors['Bearded seal'],";'>bearded,</span> ", 
                "<span style = 'color:",species_colors['Ribbon seal'],";'>ribbon,</span> ",  
                "and ", 
                "<span style = 'color:",species_colors['Spotted seal'],";'>spotted</span> seals ", 
                "across multiple regions within the Bering, Chukchi, and western Beaufort seas.")) +
  labs(caption = "shaded yellow region indicates spatial extent of aerial survey effort in US waters") +
  theme(legend.position = "none",
        axis.title = element_blank(),
        strip.background = element_blank(),
        strip.placement = "outside",
        strip.text.x = element_markdown(face = "bold"),
        plot.subtitle = element_textbox_simple(margin = ggplot2::margin(b=12)))
